### 2020-09-25 latex/vlp.tex
Minor edits to VNHC and VLP to send to Manfredi

Using this configuration, we will compute the Hamiltonian dynamics of the system.
The cartesian position of the mass at the tip of the pendulum
is given by \(x = (l\sin(q),-l\cos(q))\), while its velocity is
\(\dot{x} = (\dot{l}\sin(q) + l\cos{q}\dot{q}, -\dot{l}\cos(q) + l\sin{q}\dot{q})\).
Computing the kinetic energy \(T\) yields
\[
   T(\mathbf{q},\dot{\mathbf{q}}) = 
   \frac{1}{2}m\norm{\dot{x}}^2 = \frac{1}{2}m\left(\dot{l}^2 + l^2\dot{q}^2\right)
\]
The potential energy \(P\) with respect to the pivot (under a gravitational
acceleration \(g\)) is
\[
   P(\mathbf{q}) = -mgl\cos(q)
\]
Collecting the kinetic energy into a quadratic form, we get the Lagrangian
\[
   \mathcal{L}(\mathbf{q},\dot{\mathbf{q}}) 
   = \frac{1}{2} \dot{\mathbf{q}}\tpose D(\mathbf{q})\dot{\mathbf{q}} - P(\mathbf{q})
   = \frac{1}{2}
   \begin{bmatrix} \dot{q} & \dot{l} \end{bmatrix}
   \begin{bmatrix}
      ml^2 & 0 \\
      0 & m \\
   \end{bmatrix}
   \begin{bmatrix} 
      \dot{q} \\ \dot{l}
   \end{bmatrix}
   + mgl\cos(q)
\]
Computing the conjugate of momenta to \(\mathbf{q}\), we get 
\[
   \mathbf{p} := \begin{bmatrix} p \\ p_l \end{bmatrix} 
   = \begin{bmatrix} ml^2\dot{q} \\ m\dot{l} \end{bmatrix} 
\]
Performing the Legendre transform on \(\mathcal{L}\) and setting
\(M(\mathbf{q}) := D(\mathbf{q})\), \(V(\mathbf{q}) := P(\mathbf{q})\),
we find the Hamiltonian is equal to the total mechanical energy of the system:
\[
   \mathcal{H} = E = \frac{1}{2}\mathbf{p}\tpose \Minv(\mathbf{q})
   \mathbf{p}+V(\mathbf{q})
\]
Taking the appropriate derivatives, the dynamics of the VLP in Hamiltonian form
are described in (\ref{eqn:vlp-hamiltonian-with-pl}). 
\begin{align}\label{eqn:vlp-hamiltonian-with-pl}
   \mathcal{H} &= \frac{1}{2} \begin{bmatrix} p & p_l \end{bmatrix}
      \begin{bmatrix}
         \frac{1}{ml^2}  & 0 \\
         0 & \frac{1}{m}
      \end{bmatrix} \begin{bmatrix} p \\ p_l \end{bmatrix} - mgl\cos(q) \\
     &\begin{cases}
        \dot{q} = \frac{p}{ml^2} \\
        \dot{l} = \frac{p_l}{m} \\
        \dot{p} = -mgl\sin(q) \\
        \dot{p}_l = \frac{p^2}{ml^3} + mg\cos(q) + \tau \\
   \end{cases} \nonumber
\end{align}
The control input is a force \(\tau \in \mathbb{R}\) affecting the dynamics of
\(p_l\), acting colinearly with direction of the rod.
We assume the force does not affect the dynamics of \(p\) in any way -
that is, the control input cannot enact any lateral force on the pendulum.
This is extremely useful for simplifying the dynamics: since one can choose
\(\tau\) which sets \(\dot{p}_l\) to any desired function, we can assume (with abuse
of notation) that \(l\) is tracking some known function \(l(t)\). 
The closed loop dynamics of the system will be described exclusively by
\((\dot{q},\dot{p})\) with \(l \equiv l(t)\).
The fact that \(p_l\) does not appear in these closed-loop dynamics allows us to
ignore the subdynamics \((\dot{l},\dot{p}_l)\) entirely.

@@

A child's optimal pumping strategy is straightforward:
they stand at the lowest point of the swing, and squat at the highest point.
Looking at the VLP representation, the pendulum shortens at the bottom of the
swing, and lengthens at the top. 
For an intuitive explanation, conservation of angular momentum indicates that
shortening the pendulum at the bottom forces the mass to gain speed to
compensate for the reduced length \cite{how_to_pump_a_swing}.
Energy is not conserved in this process, so the pendulum gains kinetic energy
and reaches a higher point at the peak of its swing.
Lengthening the pendulum when it reaches this peak means gravity
imparts a larger angular momentum to the mass by the time it reaches the bottom
of its swing, which in turn is converted to a higher velocity when the
pendulum is shortened.
By alternating these processes, the pendulum experiences an average net gain in
rotational energy.

@@

We now define a continuous function which
approximates \(l^\star(\theta)\) by attaching sinusoids at the
discontinuities (see Figure \ref{fig:vlp-T-controller}). 
Let \(\Delta l := (\overline{l} - \underline{l})/2\) and 
\(l_{\text{avg}} := (\overline{l} + \underline{l})/2\).
Providing these newly attached sinusoids with a frequency \(\omega = \frac{\pi}{T}\) 
(where \(T \in \, ]0,\frac{\pi}{2}]\) is a control parameter), we get the
twice-differentiable controller \(l_T(\theta)\) (\ref{eqn:vlp-T-controller}).
\begin{equation}\label{eqn:vlp-T-controller}
   l_T(\theta) = \begin{cases}
      \overline{l} & \theta \in \left[-\frac{\pi}{2} + \frac{T}{2}, -\frac{T}{2}\right] 
      \cup \left[\frac{\pi}{2} + \frac{T}{2}, \pi - \frac{T}{2}\right] \\
      \underline{l} & \theta \in \left[-\pi + \frac{T}{2}, -\frac{\pi}{2} - \frac{T}{2}\right] 
      \cup \left[\frac{T}{2}, \frac{\pi}{2} - \frac{T}{2}\right] \\
      -\Delta l \sin(\omega(\theta + \pi)) + l_{\text{avg}} & \theta \in
      \left[-\pi,-\pi + \frac{T}{2}\right] \\
      -\Delta l \sin(\omega \theta) + l_\text{avg} & \theta \in [-\frac{T}{2},
      \frac{T}{2}] \\
      \Delta l \sin(\omega(\theta - a)) + l_\text{avg} & 
      \theta \in \left[a - \frac{T}{2}, a + \frac{T}{2}\right] \text{ for } 
      a \in \left\{-\frac{\pi}{2}, \frac{\pi}{2}\right\} \\
      -\Delta l \sin(\omega(\theta-\pi)) & \theta \in \left[\pi - \frac{T}{2},\pi\right] \\
   \end{cases}
\end{equation}

@@

This controller approximates the optimal controller because 
\[
   \lim\limits_{T \rightarrow 0} l_T(\theta) = l^\star(\theta)
\]
Unfortunately, while \(l_T(\theta)\) is twice-differentiable, its second
derivative is only piecewise continuous for most values of \(T\).
If we wish to use it as a VNHC, we must ensure that either the generalized
forces \(\tau\) acting on \(p_l\) can be discontinuous (which is certainly
achievable by humans), or we must find a value of \(T\) where this controller is
smooth.
Thankfully, \(l_{\frac{\pi}{2}}(\theta)\) is smooth as it can be simplified into
the expression (\ref{eqn:vlp-smoothed-controller}). 
It is plotted for demonstration in Figure \ref{fig:vlp-smoothed-controller}.
\begin{equation}\label{eqn:vlp-smoothed-controller}
   l_\frac{\pi}{2}(\theta) = -\Delta l \sin(2\theta) + l_{\text{avg}}
\end{equation}

@@

Because \(l_\frac{\pi}{2}(\theta)\) is smooth and it approximates
\(l^\star(\theta)\), we set our VNHC to be
\[
   h(q,p) = l - l_\frac{\pi}{2}\left(\theta(q,p)\right)
\]
We still need to prove this VNHC will inject energy into a VLP. 
We will do so by showing the dynamics (\ref{eqn:vlp-hamiltonian}) 
diverge from the origin via an ``anti-Lyapunov function".

@@

\begin{thm}\label{thm:vlp-energy-stabilization}
   Take the Hamiltonian VLP (\ref{eqn:vlp-hamiltonian}) with initial condition 
   \((q_0,p_0) \not \in \left\{(0,0),(\pm\pi,0)\right\}\).
   Define \(\theta := \arctan_2(p,q)\). 
   A VNHC of the form \(h(q,p) = l - l(\theta)\) injects energy if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l(\theta) - l_\text{avg}\right)sin(2\theta) \leq 0 \text{ }\forall \theta \in \mathbb{S}^1
   \end{equation}
   with the property that the inequality is strict for almost every \(\theta\).
   If the inequality is flipped, the VNHC dissipates energy. 
\end{thm}
\begin{proof}
    Choose, as a candidate anti-Lyapunov function, the energy for the
    average-length pendulum with zero-potential at the bottom of the swing:
    \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
                    + m g l_\text{avg} (1-\cos(q))
    \]
    This is positive definite at \((0,0)\) and has derivative 
    \[
      \dot{E}_\text{avg} = \frac{-g\sin(q)p \left(l(\theta)^3 - l_\text{avg}^3\right)}
                 {l_\text{avg}^2l(\theta)^2}
    \]
    The sign of \(\sin(q)p\) is dependent only on the quadrant in the
    \((q,p)\)-plane, so it is equal to the sign of \(\sin(2\theta)\). 
    Furthermore, by Lemma \ref{lemma:sign-of-cube}:
    \[ 
       \sign{l(\theta)^3 - l_\text{avg}^3} = \sign{ l(\theta) - l_\text{avg}}
    \]
    Thus, the derivative of \(E_\text{avg}\) is almost always positive since
    \begin{align*}
       \label{eq:}
       \sign{\dot{E}_\text{avg}} &= \sign{-\sin(q)p \left(l(\theta)^3 - l_\text{avg}^3\right)} \\
                   &= -\sign{\sin(2\theta) \left(l(\theta) - l_\text{avg}\right)} \\
                   &\geq 0 \text{ (by assumption)}
    \end{align*}
    Therefore \(E_\text{avg}\) is an anti-Lyapunov function.
    Since the initial condition is not an equilibrium,
    the orbit of the VNHC in the \((q,p)\) plane is diverging away from the
    origin and the VLP is gaining energy on average.
    Flipping the inequality of (\ref{eqn:vlp-energy-gain-condition}) 
    means \(E_\text{avg}\) is a regular Lyapunov function, so the VLP is
    losing energy while converging to \((0,0)\).
 \end{proof}

@@

All together, these theory in this chapter shows that VNHCs can replicate the
time-optimal pumping strategy performed by humans. 
Furthermore, we see that VNHCs are a powerful tool for creating simple energy
stabilization techniques based on natural human motion.

>>>>

Using this configuration, we will compute the Hamiltonian dynamics of the system.
The cartesian position of the mass at the tip of the pendulum
is given by \(x = (l\sin(q),-l\cos(q))\), while its velocity is
\(\dot{x} = (\dot{l}\sin(q) + l\cos(q)\dot{q}, -\dot{l}\cos(q) + l\sin(q)\dot{q})\).
Computing the kinetic energy \(T\) yields
\[
   T(\mathbf{q},\dot{\mathbf{q}}) = 
   \frac{1}{2}m\norm{\dot{x}}^2 = \frac{1}{2}m\left(\dot{l}^2 + l^2\dot{q}^2\right)
\]
The potential energy \(P\) with respect to the pivot (under a gravitational
acceleration \(g\)) is
\[
   P(\mathbf{q}) = -mgl\cos(q)
\]
Collecting the kinetic energy into a quadratic form, we get the Lagrangian
\[
   \mathcal{L}(\mathbf{q},\dot{\mathbf{q}}) 
   = \frac{1}{2} \dot{\mathbf{q}}\tpose D(\mathbf{q})\dot{\mathbf{q}} - P(\mathbf{q})
   = \frac{1}{2}
   \begin{bmatrix} \dot{q} & \dot{l} \end{bmatrix}
   \begin{bmatrix}
      ml^2 & 0 \\
      0 & m \\
   \end{bmatrix}
   \begin{bmatrix} 
      \dot{q} \\ \dot{l}
   \end{bmatrix}
   + mgl\cos(q)
\]
Computing the conjugate of momenta to \(\mathbf{q}\), we get 
\[
   \mathbf{p} := \begin{bmatrix} p \\ p_l \end{bmatrix} 
   = \begin{bmatrix} ml^2\dot{q} \\ m\dot{l} \end{bmatrix} 
\]
Performing the Legendre transform on \(\mathcal{L}\) and setting
\(M(\mathbf{q}) := D(\mathbf{q})\), \(V(\mathbf{q}) := P(\mathbf{q})\),
we find the Hamiltonian is equal to the total mechanical energy of the system:
\[
   \mathcal{H} = E = \frac{1}{2}\mathbf{p}\tpose \Minv(\mathbf{q})
   \mathbf{p}+V(\mathbf{q})
\]
Taking the appropriate derivatives, the dynamics of the VLP in Hamiltonian form
are described in (\ref{eqn:vlp-hamiltonian-with-pl}). 
\begin{align}\label{eqn:vlp-hamiltonian-with-pl}
   \mathcal{H} &= \frac{1}{2} \begin{bmatrix} p & p_l \end{bmatrix}
      \begin{bmatrix}
         \frac{1}{ml^2}  & 0 \\
         0 & \frac{1}{m}
      \end{bmatrix} \begin{bmatrix} p \\ p_l \end{bmatrix} - mgl\cos(q) \\
     &\begin{cases}
        \dot{q} = \frac{p}{ml^2} \\
        \dot{l} = \frac{p_l}{m} \\
        \dot{p} = -mgl\sin(q) \\
        \dot{p}_l = \frac{p^2}{ml^3} + mg\cos(q) + \tau \\
   \end{cases} \nonumber
\end{align}
The control input is a force \(\tau \in \mathbb{R}\) affecting the dynamics of
\(p_l\), acting colinearly with direction of the rod.
We assume the force does not affect the dynamics of \(p\) in any way -
that is, the control input cannot enact any lateral force on the pendulum.
This is extremely useful for simplifying the dynamics: since one can choose
\(\tau\) which sets \(\dot{p}_l\) to any desired function and \(\dot{l}\)
depends only on \(p_l\), we can assume (with abuse of notation) that \(l\) is
tracking some known function \(l(t)\). 
The closed loop dynamics of the system will be described exclusively by
\((\dot{q},\dot{p})\) with \(l \equiv l(t)\).
The fact that \(p_l\) does not appear in these closed-loop dynamics allows us to
ignore the subdynamics \((\dot{l},\dot{p}_l)\) entirely.

@@

A child's optimal pumping strategy is the following: 
they stand at the lowest point of the swing, and squat at the highest point.
Looking at the VLP representation, the pendulum shortens at the bottom of the
swing, and lengthens at the top. 
For an intuitive explanation, conservation of angular momentum indicates that
shortening the pendulum at the bottom forces the mass to gain speed to
compensate for the reduced length \cite{how_to_pump_a_swing}.
Energy is not conserved in this process, so the pendulum gains kinetic energy
and reaches a higher point at the peak of its swing.
Lengthening the pendulum when it reaches this peak means gravity
imparts a larger angular momentum to the mass by the time it reaches the bottom
of its swing, which in turn is converted to a higher velocity when the
pendulum is shortened.
By alternating these processes, the pendulum experiences an average net gain in
rotational energy.

@@

We now define a continuous function which
approximates \(l^\star(\theta)\) by attaching sinusoids at the
discontinuities (see Figure \ref{fig:vlp-T-controller}). 
Let \(\Delta l := (\overline{l} - \underline{l})/2\) and 
\(l_{\text{avg}} := (\overline{l} + \underline{l})/2\).
Providing these newly attached sinusoids with a frequency \(\omega = \frac{\pi}{T}\) 
(where \(T \in \, ]0,\frac{\pi}{2}]\) is a control parameter), we get the
continuously-differentiable controller \(l_T(\theta)\) (\ref{eqn:vlp-T-controller}).
\begin{equation}\label{eqn:vlp-T-controller}
   l_T(\theta) = \begin{cases}
      \overline{l} & \theta \in \left[-\frac{\pi}{2} + \frac{T}{2}, -\frac{T}{2}\right] 
      \cup \left[\frac{\pi}{2} + \frac{T}{2}, \pi - \frac{T}{2}\right] \\
      \underline{l} & \theta \in \left[-\pi + \frac{T}{2}, -\frac{\pi}{2} - \frac{T}{2}\right] 
      \cup \left[\frac{T}{2}, \frac{\pi}{2} - \frac{T}{2}\right] \\
      -\Delta l \sin(\omega(\theta + \pi)) + l_{\text{avg}} & \theta \in
      \left[-\pi,-\pi + \frac{T}{2}\right] \\
      -\Delta l \sin(\omega \theta) + l_\text{avg} & \theta \in [-\frac{T}{2},
      \frac{T}{2}] \\
      \Delta l \sin(\omega(\theta - a)) + l_\text{avg} & 
      \theta \in \left[a - \frac{T}{2}, a + \frac{T}{2}\right] \text{ for } 
      a \in \left\{-\frac{\pi}{2}, \frac{\pi}{2}\right\} \\
      -\Delta l \sin(\omega(\theta-\pi)) & \theta \in \left[\pi - \frac{T}{2},\pi\right] \\
   \end{cases}
\end{equation}

@@

This controller approximates the optimal controller because 
\[
   \lim\limits_{T \rightarrow 0} l_T(\theta) = l^\star(\theta)
\]
Unfortunately, while \(l_T(\theta)\) is continuously-differentiable, 
it is not twice-differentiable for most values of \(T\).
If we wish to use it as a VNHC, we must ensure that either the generalized
forces \(\tau\) acting on \(p_l\) can be discontinuous (which is certainly
achievable by humans), or we must find a value of \(T\) where this controller is
smooth.
Thankfully, \(l_{\frac{\pi}{2}}(\theta)\) is smooth as it can be simplified into
the expression (\ref{eqn:vlp-smoothed-controller}). 
\begin{equation}\label{eqn:vlp-smoothed-controller}
   l_\frac{\pi}{2}(\theta) = -\Delta l \sin(2\theta) + l_{\text{avg}}
\end{equation}
It is plotted for demonstration in Figure \ref{fig:vlp-smoothed-controller}.

@@

Because \(l_\frac{\pi}{2}(\theta)\) is smooth and it approximates
\(l^\star(\theta)\), we set our VNHC to be
\[
   h(q,p) = l - l_\frac{\pi}{2}\left(\theta(q,p)\right)
\]
We still need to prove this VNHC will inject energy into a VLP. 
We will do so by showing the dynamics (\ref{eqn:vlp-hamiltonian}) 
diverge from the origin via a so-called anti-Lyapunov function.

@@

\begin{thm}\label{thm:vlp-energy-stabilization}
   Take the VLP (\ref{eqn:vlp-hamiltonian}) with initial condition 
   \((q_0,p_0) \not \in \left\{(0,0),(\pm\pi,0)\right\}\).
   Define \(\theta := \arctan_2(p,q)\). 
   A VNHC of the form \(h(q,p) = l - l(\theta)\) injects energy if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l(\theta) - l_\text{avg}\right)sin(2\theta) \leq 0 \text{ }\forall \theta \in \mathbb{S}^1
   \end{equation}
   with the property that the inequality is strict for almost every \(\theta\).
   If the inequality is flipped, the VNHC dissipates energy. 
\end{thm}
\begin{proof}
    Choose, as a candidate anti-Lyapunov function, the energy for the
    average-length pendulum with zero-potential at the bottom of the swing:
    \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
                    + m g l_\text{avg} (1-\cos(q))
    \]
    This is positive definite at \((0,0)\) and has derivative 
    \[
      \dot{E}_\text{avg} = \frac{-g\sin(q)p \left(l(\theta)^3 - l_\text{avg}^3\right)}
                 {l_\text{avg}^2l(\theta)^2}
    \]
    The sign of \(\sin(q)p\) is dependent only on the quadrant in the
    \((q,p)\)-plane, so it is equal to the sign of \(\sin(2\theta)\). 
    Furthermore, by Lemma \ref{lemma:sign-of-cube}:
    \[ 
       \sign{l(\theta)^3 - l_\text{avg}^3} = \sign{ l(\theta) - l_\text{avg}}
    \]
    Thus, the derivative of \(E_\text{avg}\) is almost always positive since
    \begin{align*}
       \label{eq:}
       \sign{\dot{E}_\text{avg}} &= \sign{-\sin(q)p \left(l(\theta)^3 - l_\text{avg}^3\right)} \\
                   &= -\sign{\sin(2\theta) \left(l(\theta) - l_\text{avg}\right)} \\
                   &\geq 0 \text{ (by assumption)}
    \end{align*}
    Therefore \(E_\text{avg}\) is an anti-Lyapunov function.
    Since the initial condition is not an equilibrium,
    the orbit of the VNHC in the \((q,p)\) plane is diverging away from the
    origin and the VLP is gaining energy on average.
    Flipping the inequality of (\ref{eqn:vlp-energy-gain-condition}) 
    means \(E_\text{avg}\) is a regular Lyapunov function, so the VLP is
    losing energy while converging to \((0,0)\).
 \end{proof}

@@

All together, the theory developed in this chapter shows that VNHCs can
replicate the time-optimal pumping strategy performed by humans. 
Furthermore, we see that VNHCs are a powerful tool for creating simple energy
stabilization techniques based on natural human motion.

### 2020-09-25 latex/vnhcs.tex
Added a question for Manfredi
Minor edits to VNHC and VLP to send to Manfredi

\begin{figure}
   \centering
   \begin{subfigure}[]{0.75\textwidth}
      \includestandalone[width=\textwidth]{images/newtonian_barbell}
      \caption{The Newtonian representation of the barbell 
      requires all six cartesian positions and the corresponding EOC.}
   \end{subfigure}
   \begin{subfigure}[]{0.75\textwidth}
      \includestandalone[width=\textwidth]{images/gc_barbell}
      \caption{One possible set of three generalized coordinates is
       \((x_0,y_0,\theta)\), which represent the position of the 
       center of the bar and the angle of the barbell in the \(xy\)-plane.}
   \end{subfigure}
   \caption{A mechanical system with \(N = 2\) points at \((x_1,y_1,z_1)\)
   and \((x_2,y_2,z_2)\) separated by a bar with \(r = 3\) EOC given by 
   \(c = \norm{x_1 - x_2}\), \(z_1 = 0\), and \(z_2 = 0\). This system has \(n =
   3\) degrees of freedom.}
   \label{fig:barbell}
\end{figure}

@@

% Perform the full development of VNHCs, including its stabilizing
% controller.
\begin{defn}
    A \textit{virtual nonholonomic constraint} (VNHC) \textit{of order \(k\)} is a
    relation \(h(q,p) = 0\) where \(h : \mathcal{Q}\times\R^n \rightarrow \R^k\) is
    smooth, \(\rank{\left[ dh_q,\, dh_p \right]} = k\) for all 
    \((q,p) \in h\inv(0)\), and there exists a feedback controller \(\tau(q,p)\)
    stabilizing the set
    \[
        \Gamma = \left\{(q,p) \mid h(q,p) = 0, dh_q \dot{q} + dh_p \dot{p} = 0\right\}
    \]
    which is called the \textit{constraint manifold}.
\end{defn}

@@

    That is, a regular VNHC is a function 
    \(h : \mathcal{Q} \times \mathcal{P}_u \rightarrow \R^k\) which satisfies
    the rank condition.

@@

\begin{cor}\label{cor:2d-zero-dynamics}
    Suppose \(\mathcal{H}\) is a ---- mechanical system satisfying Assumption
    \ref{assm:H-is-simply-actuated} which has degree of underactuation one.
    Let \(h(q,p_u) = 0\) be a regular VNHC of order \((n-1)\) of the form
    \(h(q,p_u) = q_a - f(q_u,p_u)\)
    where \(f\) is a suitably-defined smooth function.
    Then \(dh_q = \begin{bmatrix} -\partial f_{q_u} & \Id{(n-1)}
    \end{bmatrix}\).
    Defining \(e_1 := (1,0,\ldots,0) \in \R^n\), the actuated momentum is given
    by (\ref{eqn:g-qupu}):
    \begin{equation}\label{eqn:g-qupu}
        p_a = -\left.\left(dh_q \Minv(q)
        \begin{bmatrix}
            \Zmat{1\times (n-1)}\\
            I_{(n-1)}
        \end{bmatrix}\right)\inv 
        \left(\partial_{p_u}f \partial_{q_u}V + dh_q \Minv(q)e_1 p_u\right) 
            \right|_{q_a =f(q_u,p_u)}
    \end{equation}
    Since \(q_u \in \Rt{T}\) for some \(T \in ]0,\infty]\) and \(p_u \in \R\),
    the orbit \((q_u(t),p_u(t))\) traces out
    a curve on the 2D-plane \(\Rt{T} \times \R\) which we call the \((q,p)\)-plane.
\end{cor}

>>>>

\begin{figure}
   \centering
   \begin{subfigure}[]{0.75\textwidth}
      \includestandalone[width=\textwidth]{images/newtonian_barbell}
      \caption{The Newtonian representation of the barbell 
      requires all six cartesian positions and the corresponding EOC.}
   \end{subfigure}
   \begin{subfigure}[]{0.75\textwidth}
      \includestandalone[width=\textwidth]{images/gc_barbell}
      \caption{One possible set of three generalized coordinates is
       \((x_0,y_0,\theta)\), which represent the position of the 
       center of the bar and the angle of the barbell in the \(xy\)-plane.}
   \end{subfigure}
   \caption{A mechanical system with \(N = 2\) point masses at \(a = (x_1,y_1,z_1)\)
       and \(b = (x_2,y_2,z_2)\), separated by a bar of length \(c\). 
       There are \(r = 3\) EOC given by \(\norm{a - b} = c\), \(z_1 = 0\),
       and \(z_2 = 0\). This system has \(n = 3\) degrees of freedom.}
   \label{fig:barbell}
\end{figure}

@@

% Perform the full development of VNHCs, including its stabilizing
% controller.
\begin{defn}
    A \textit{virtual nonholonomic constraint} (VNHC) \textit{of order \(k\)} is a
    relation \(h(q,p) = 0\) where \(h : \mathcal{Q}\times\R^n \rightarrow \R^k\) is
    smooth, \(\rank{\left[ dh_q,\, dh_p \right]} = k\) for all 
    \((q,p) \in h\inv(0)\), and there exists a feedback controller \(\tau(q,p)\)
    stabilizing the set
    \[
        \Gamma = \left\{(q,p) \mid h(q,p) = 0, dh_q \dot{q} + dh_p \dot{p} = 0\right\}
    \]
    which is called the \textit{constraint manifold}.
\end{defn}
\textbf{Manfredi: Do we require \(C^\infty\), or can we relax it to \(C^2\)?}

@@

@@

\begin{cor}\label{cor:2d-zero-dynamics}
    Suppose \(\mathcal{H}\) is a ---- mechanical system satisfying Assumption
    \ref{assm:H-is-simply-actuated} which has degree of underactuation one.
    Let \(h(q,p_u) = 0\) be a regular VNHC of order \((n-1)\) of the form
    \(h(q,p_u) = q_a - f(q_u,p_u)\)
    where \(f\) is a suitably-defined smooth function.
    Then \(dh_q = \begin{bmatrix} -\partial_{q_u}f & \Id{(n-1)}
    \end{bmatrix}\).
    Defining \(e_1 := (1,0,\ldots,0) \in \R^n\), the actuated momentum is given
    by (\ref{eqn:g-qupu}):
    \begin{equation}\label{eqn:g-qupu}
        p_a = -\left.\left(dh_q \Minv(q)
        \begin{bmatrix}
            \Zmat{1\times (n-1)}\\
            I_{(n-1)}
        \end{bmatrix}\right)\inv 
        \left(\partial_{p_u}f \partial_{q_u}V + dh_q \Minv(q)e_1 p_u\right) 
            \right|_{q_a =f(q_u,p_u)}
    \end{equation}
    Since \(q_u \in \Rt{T}\) for some \(T \in ]0,\infty]\) and \(p_u \in \R\),
    the orbit \((q_u(t),p_u(t))\) traces out
    a curve on the 2D-plane \(\Rt{T} \times \R\) which we call the \((q,p)\)-plane.
\end{cor}
